---
title: "Markdown File"
author: "Hannah Damico"
date: "`r Sys.Date()`"
output:
  html_document: default
---

```{r setup, include=FALSE}
# the line below set this as your global chunk
# anything you run here will apply to all chunks in your markdown
# this is the best place to load a library
# echo = TRUE means when we knit this document, the code will be visible
knitr::opts_chunk$set(echo = TRUE)
# install a package
# install.packages("tidyverse")
# load a library
library(tidyverse)
# these libraries below are contained within tidyverse
# so loading them like this can be redundant, but not harmful
library(dplyr)
library(ggplot2)
?tidyverse

search() # what packages do we have uploaded right now?
```


 <font size="4"> 



# HTML Document Title {.tabset}

This isn't a tab, but a secondary document title.

You can run a command a few ways

Mac: Command + Return     
PC: Ctl + Enter     

Insert a code chunk: 
Mac: Command + Option + i   
PC: Ctl + Alt + i


**Double asterisk makes text bold**   
__So does double underscore__   
*Single asterisks make things italicized*    
_So does single underscore_      
   

```{r}
# control + option + i to create a code chunk 
# Mac users: command + return to run code line that cursor is on
# PC users: control + enter to run code line that cursor is on
# comment 1
```


## Help Functions + R Studio Interface - First Tab Title



```{r}
# need to understand your library or function more?
?select
# want to specify where your function is coming from?
?dplyr::select
# why is this useful?
# load in this package 
library(MASS)
# try loading ?select again
?select
# now R studio is asking which version of select you want to look at
?dplyr::select()
?MASS::select()

# the order that packages are loaded in matters
# MASS was loaded second, so it's select function will be the default

# What functions are present in dplyr?
# lsf.str("package:dplyr")
```



## Create an R Object - Second Tab Title

### Types of Objects

**Logical**   
Useful to check the type of object we have   
TRUE or FALSE
Sometimes 0 or 1

**Character**    
Objects in quotations such as: "A", "B", "1", "2" or 'a', 'b', '1', '2'  
Single or double quotes work, but should be consistent!

**String**     
A group of characters together in quotes such as: "sAAg", "aBc", "ginko", "KuaiKuai", "t0asta0", "nongSHIM"    

**Factors**

**Numeric**      
A real number (negative or positive numbers), fractions, or decimals    


**Vectors**   
Two types:   

1. Atomic vectors: Elements are consistent object types, e.g. numerics only or characters only            
2. List vectors: Elements can be a combination of object types, e.g. numerics + characters         


&nbsp;
&nbsp;
&nbsp;
&nbsp;


#### How do we save things?
Use:     
Arrow: <-    
Equal sign: =      

Consistency is important for reproducibility and clean code!


```{r}
?class # what is this?
?typeof # WHAT IS THIS??
?storage.mode() # WHAT IN THE WORLD IS THIS???
```



```{r}
# Logic 
# DONT SAVE THINGS TO THE NAMES TRUE AND FALSE
# We don't see the below example much
A <- TRUE
B <- FALSE
# Character
char <- "A"
is.character(char) # are we sure this is a character? Return TRUE or FALSE
# String
string1 <- "beans"
is.character(string1) # Will this be true or false?
nchar(string1) # how long is our string?
# Numeric
num1 <- 1
num2 <- "2" # Is this a numeric?
is.numeric(num1)
is.numeric(num2)
num3 <- as.numeric(num2) # How about now? What will this return? num2 <- as.numeric(num2)

```

Non-breaking space! 

&nbsp;
&nbsp;

```{r}
# Vectors
vec1 <- c("a", "b", "c")
vec2 <- c(2,4,5,678,9,10000)
# So what does c() do?
# Concatenate! Link things together 
# What happens when we link these vectors together?
c(vec1, vec2)
```


```{r}
# Matrix
# Why do we get this warning?
matrix(vec1, nrow = 3)
# now we have 3 columns and 1 row
matrix(vec1, ncol = 3)
# What happens if we give it two instructions, but we run out of vector elements?
matrix(vec1, ncol = 3, nrow= 2)
matrix(vec1, nrow = 3, ncol= 2)

# Let's use that second matrix example - we have to save it
mat <- matrix(vec1, nrow = 3, ncol= 2)
# let's turn it into something new

# Dataframe
as.data.frame(mat)

# Let do this another way
data.frame(vec1,vec1)
# Notice how the vectors save as columns this way
# We can also give our columns names!
data.frame(letters1 = vec1, letters2 = vec1)
# Or this way, but note that we must save the dataframe first
df <- data.frame(vec1,vec1)
colnames(df) <- c("letters1", "letters2")
# Notice how we used a vector or strings
```


&nbsp;
&nbsp;

## Viewing your Data - Third Tab Title


```{r}
# where do we read it from?
?read.csv()
?read_csv()
# our working directory!
getwd() # where is the homebase, the workshop for where our markdown is stored?
```

## Analysis 

```{r}
library(readr) # this is in tidyverse!
# Is anything wrong with this?
#cytomegalovirus <- read_csv("~/bbc-research/RCR/R-Workshop-Sept2024/cytomegalovirus.csv")
# View(cytomegalovirus) # note that the document will not knit unless this is commented out 

cyto_data <- read_csv("Data/cytomegalovirus.csv")

# read.csv("Data/cytomegalovirus.csv")

```


```{r}
# Understand our data more
## What kind of object is our data?
class(cyto_data)
## What are the dimensions of our data? How many rows, how many columns
dim(cyto_data) # 64 rows, 26 columns
# We can also use 
nrow(cyto_data) # to only count rows
ncol(cyto_data) # to only count columns
# Let's take a glimpse at our data
glimpse(cyto_data)
# Notice that we can see the variable types
# dbl = double and is a number that can include a decimal
# Here's another option
str(cyto_data)

# Let's  print out our data this way
cyto_data
# try running this in your console and see what happens!
```


```{r}
# Let's look at the first 10 rows of our data
head(cyto_data, n= 10)

# what about the last 10 rows?
tail(cyto_data, n = 10)
```


&nbsp;
&nbsp;




</font> 




























